Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-05-04 second-draft/pnas-word-order-si.tex
si

We base our formalization on the comprehensive study by Dryer \cite{dryer1992greenbergian}.\footnote{Regarding recent objections by~\cite{dunn2011evolved}, see follow-ups by \cite{levy2011computational}, and \cite{croft2011greenbergian}.}
Greenberg's original study was based on 30 languages; more recently, Dryer \cite{dryer1992greenbergian} documented the word order correlations based on typological data from 625 languages.
\cite{dryer1992greenbergian} formulated these universals as correlations between the order of objects and verbs and the orders of other syntactic relations.
We test our ordering grammars for these correlations by testing whether the coefficients for these syntactic relations have the same sign as the coefficient of the verb-object relation.
Testing correlations is therefore constrained by the degree to which these relations are annotated in UD.
The verb-object relation corresponds to the  \emph{obj} relation defined by UD.
While most of the other relations also correspond to UD relations, some are not annotated reliably.
We were able formalize eleven out of Dryer's seventeen correlations in UD. Six of these could not be expressed individually in UD, and were collapsed into three coarse-grained correlations.

We excluded four correlations that are not annotated reliably in UD, and are only relevant to some of the world's languages: Question particles, plural words (i.e., independent plural markers), and articles.
All three types of elements occur at most in parts of the 51 UD languages, and none of them is annotated reliably in those languages where they occur.
Among these three types of elements, the one most prominent in our sample of 51 languages is articles (occurring in many European languages); UD subsumes them under the \emph{det} relation, which is also used for other frequent elements, such as demonstratives and quantifiers.

@@

We further excluded the verb-subject correlation, which is not satisfied by much more than half of the world's languages (51 \% among those with annotation in \cite{wals-81}, with clear violation in 35.4 \%).
It is satisfied only in 33\% of our sample of 51 UD languages, as quantified using the grammars we extracted.
Dryer \cite{dryer1992greenbergian} counts this as a correlation since he describes the distribution of subject order as an interaction between a weak correlation with object order, and a very strong dominance principle favoring SV/SO orderings.
We focus on the modeling of correlations, and such dominance principles are beyond the scope of our experiments.
We therefore excluded this correlation.
%We note that efficiency makes predictions compatible with the pattern observed in our sample of 51 languages, potentially in line with the noisy-channel explanation of SVO described by~\cite{gibson2013noisy} (see Table~\ref{tab:all-predictions}).

TODO collapsing

@@

\begin{table*}[ht]
	\begin{center}
\begin{tabular}{|ll|l|l}
	\hline
	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{Operationalization}       \\ 
	verb & object &    \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP    &    lifted\_case   & G3, G4   \\ \hline
copula  verb  &    predicate    &    lifted\_cop   & --    \\\hline
`want'    &    VP    &    xcomp   & G15   \\\hline
tense/aspect auxiliary    &    VP    &    \multirow{2}{*}{aux}   & G16, G13  \\
negative auxiliary    &    VP    &     \\ \hline
complementizer    &    S    &   \multirow{2}{*}{lifted\_mark}  & --    \\
adverbial subordinator & S & \\ \hline
noun    &    genitive    &    nmod & G2, G23   \\ \hline
noun    &    relative clause    &    acl   &  G24    \\ \hline
adjective & standard of comparison & \multirow{2}{*}{obl}\\
verb    &    PP    &    & G22   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{3}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations according to Dryer \cite{dryer1992greenbergian}, with operationalizations with Universal Dependencies. We also provide the corresponding numbers in Greenberg's original study.}\label{table:greenberg-dryer}
\end{table*}

@@

   	\caption{.}
    \label{fig:todo}
\end{figure}

@@

Here, we validate our grammars by comparing the dominant orders of six syntactic relations that are also annotated in WALS.

WALS has data for 74\% of the entries, and lists a dominant order for 91\% of these.
96~\% of entries where WALS records a dominant word order agree with the WALS judgment.

$*$ indicates that WALS does not list a dominant order, ? indicates that WALS has no data, and $--$ indicates that the relation does not occur in the corpus.

>>>>

We base our formalization on the comprehensive study by Dryer \cite{dryer1992greenbergian}.\footnote{Regarding recent objections by~\cite{dunn2011evolved}, see follow-ups by \cite{levy2011computational}, and \cite{croft2011greenbergian}.}
Greenberg's original study was based on 30 languages; more recently, Dryer \cite{dryer1992greenbergian} documented the word order correlations based on typological data from 625 languages.
\cite{dryer1992greenbergian} formulated these universals as correlations between the order of objects and verbs and the orders of other syntactic relations.
We test our ordering grammars for these correlations by testing whether the coefficients for these syntactic relations have the same sign as the coefficient of the verb-object relation.
Testing correlations is therefore constrained by the degree to which these relations are annotated in UD.
The verb-object relation corresponds to the  \emph{obj} relation defined by UD.
While most of the other relations also correspond to UD relations, some are not annotated reliably.
We were able formalize eleven out of Dryer's seventeen correlations in UD.
Six of these could not be expressed individually in UD, and were collapsed into three coarse-grained correlations:
First, tense/aspect and negative auxiliaries are together represented by the \emph{aux} relation in UD.
Second, the relation between complementizers and adverbial subordinators with their complement clauses is represented by the \emph{lifted\_mark} relation.
Third, we both the verb-PP relation and the relation between adjectives and their standard of comparison is captured by the \emph{obl} relation.

We excluded three correlations that are not annotated reliably in UD, and are only relevant to some of the world's languages: Question particles, plural words (i.e., independent plural markers), and articles.
All three types of elements occur at most in parts of the 51 UD languages, and none of them is annotated reliably in those languages where they occur.
Among these three types of elements, the one most prominent in our sample of 51 languages is articles (occurring in many European languages); UD subsumes them under the \emph{det} relation, which is also used for other frequent elements, such as demonstratives and quantifiers.

@@

We further excluded the verb-subject correlation, which is not satisfied by much more than half of the world's languages (51 \% among those with annotation in \cite{wals-81}, with clear violation in 35.4 \%).
It is satisfied only in 33\% of our sample of 51 UD languages, as quantified using the grammars we extracted.
Dryer \cite{dryer1992greenbergian} counts this as a correlation since he describes the distribution of subject order as an interaction between a weak correlation with object order, and a very strong dominance principle favoring SV orderings.
We focus on the modeling of correlations, and leave dominance principles to future research.
We therefore excluded this correlation here.
%We note that efficiency makes predictions compatible with the pattern observed in our sample of 51 languages, potentially in line with the noisy-channel explanation of SVO described by~\cite{gibson2013noisy} (see Table~\ref{tab:all-predictions}).

@@

\begin{table*}[ht]
	\begin{center}
\begin{tabular}{|ll|l|l|}
	\hline
	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{Operationalization}   & \multirow{2}{*}{\cite{greenberg1963universals}}    \\ 
	verb & object &    \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP    &    case   & 3, 4   \\ \hline
copula  verb  &    predicate    &    cop   & --    \\\hline
`want'    &    VP    &    xcomp   & 15   \\\hline
tense/aspect auxiliary    &    VP    &    \multirow{2}{*}{aux}   & 16, 13  \\
	negative auxiliary    &    VP    &    & -- \\ \hline
complementizer    &    S    &   \multirow{2}{*}{mark}  & --    \\
	adverbial subordinator & S &  & -- \\ \hline
noun    &    genitive    &    nmod & 2, 23   \\ \hline
noun    &    relative clause    &    acl   &  24    \\ \hline
	adjective & standard of comparison & \multirow{2}{*}{obl} & --\\
verb    &    PP    &    & 22   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{3}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations according to Dryer \cite{dryer1992greenbergian}, with operationalizations with Universal Dependencies. We also provide the corresponding numbers in Greenberg's original study.}\label{table:greenberg-dryer}
\end{table*}

@@

	\caption{Correlation between the number of satisfied correlations ($x$-axis) and efficiency, parseability, and predictability ($y$-axis), for the 51 real languages.}
    \label{fig:todo}
\end{figure}

@@

Here, we validate our grammars by comparing the dominant orders of six syntactic relations that are also annotated in WALS.
Among the eight relations that we were able to test, five are annotated in WALS.
Here, we compare our grammars with WALS on these five relations, and the verb-object relation.

WALS has some data for all but seven languages.\footnote{Note that Serbian and Croatian are listed as a single language Serbian-Croatian in WALS. In the table, we compare those with the grammar we extracted for Croatian, noting that it agrees with the Serbian grammar.}
WALS has data for 74\% of the entries, and lists a dominant order for 91\% of these.
The grammars we extracted from the corpora agree with WALS in 96~\% out of those cases where WALS records a dominant word order.

For each of the six syntactic relation, the first column provides the ordered coded in the extracted grammar; the second column provides the order coded in WALS (DH for dependent-head, HD for head-dependent order). `?' indicates that WALS has no data.
$*$ indicates that WALS does not list a dominant order, which can mean that both orders occur in the language, or that insufficient data was available when compiling WALS \citep{dryer2011evidence}.
Finally, `--' indicates that the relation does not occur in the corpus.
